Liveness-Based Garbage Collection for Lazy Languages

### 2015-07-10 paper/paper.tex
Review up to Section 4.1 over.

Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\SRETURN}.
The  expression $(\SRETURN~\psi:x)$  with a  demand $\sigma$  forces an
evaluation  of   $x$  and  hence  a  new   map  $[\mathit{\epath}  \mapsto
  \{x.\sigma\}] $  needs to  be created, where the $\epath$ is a fresh
evaluation path terminating in $\psi$.  The  map for  the expression
$(\SIF~\psi:x~e_1~e_2)$ is a pointwise-union  of the maps of $e_1$ and
$e_2$. In addition, since the condition also triggers an evaluation of
$x$,  the map  $[\mathit{\epath} \mapsto  \{x.\epsilon\}]$ is  created and
added  to the  union.  Also  notice a  consequences  of laziness---the
entire  expression including  the condition  is not  evaluated  if the
demand on it is $\emptyset$.  This results in the empty map, which, by
abuse of notation, is also denoted by $\emptyset_{\cal{M}}$.

 The liveness  of {\LET} has  an elegant formulation.   Recollect that
 our language permits lazy {\LET}s whose definitions can be recursive,
 i.e.  in the expression $\LET~x \leftarrow~s~\IN~e$, $x$ can occur in
 $s$.   We  model  the  definition  $x  \leftarrow~s$  as  a  possibly
 recursive function (named $\mathit{f}$ in the rule) which captures an
 arbitrary number of ``unrollings'' of the definition\footnote{The $*$
   in  the definition  of ${\mathit  f}$  represents non-deterministic
   choice.  During   liveness  it  can  be  assumed   to  generate  no
   liveness.}.   We pass to  the demand  transformer $\Lfonly_{\mathit
   f}$ of this function the demand  arising out of the liveness of $x$
 at the beginning of $e$. This  gives the liveness of the variables of
 $s$.

@@

We make three observations: firstly the rule ({\sc live-define}) has a
least  solution  as  $\mathcal{L}(\cdot)$  is monotonic  in  $\sigma$;
secondly  that  ({\sc  live-define})   resembles  the  rule  for  type
inference of mutually recursive  function definitions, and thirdly the
asymmetry of  demand and liveness (compared to  post- and pre-liveness
classically) is due to the functional formulation here.
}%mycomment

@@

The  solution  is  to   also  treat  the  operations  mentioned  above
symbolically.   We  introduce three  new  symbols.  These symbols  are
defined as a relation  $\hookrightarrow$ between sets of access paths.
First, \bcar\ and \bcdr\ are defined as:

\begin{align*}
  &\sigma_1\bcar\sigma_2  \hookrightarrow
  \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \acar\sigma_2',\\
&\sigma_1\bcdr\sigma_2  \hookrightarrow
  \sigma_1\sigma_2'\mbox{ where } \sigma_2 =
  \acdr\sigma_2'
\end{align*}
Thus \bcar, for example, removes the leading \acar\ from $\sigma_2$.
Similarly, the symbol \clazy\ reduces the string following it to an
$\epsilon$. \clazy\  is defined by addition to
$\hookrightarrow$ the rule:
\begin{align*}
  &\sigma_1\clazy\sigma_2  \hookrightarrow
  \sigma_1 \mbox{ if } \sigma_2 \subseteq  \{\acar,\acdr\}^\ast
\end{align*}
%% \begin{align*}
%%  \bcar\sigma &\triangleq \{\alpha \mid 0\alpha \in \sigma\}\\
%%  \bcdr\sigma &\triangleq \{\alpha \mid 1\alpha \in \sigma\}
%% \end{align*}
%% and \clazy\ as:
%% \begin{align*}
%%  \clazy\sigma \triangleq\; & \emptyset~\mbox{if}~\sigma = \emptyset\\
%%                        & \{\epsilon\} ~\mbox{otherwise}
%% \end{align*}
%% Note that \clazy\ is idempotent.
We can  now rewrite the \CONS\  and the \CAR\  rules of $\mathit{ref}$
as:
\begin{align*}
&\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
= x.\bcar\sigma \cup y.\bcdr\sigma  \label{eqn:mod-cons},~
\mbox{and} \\
&\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          =   x.\clazy\sigma \cup x.0\sigma
\end{align*}
and the \Lfunonly\ rule
for \SIF\ as:
\begin{align*}
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) =
                    &\mathcal{L}(e_1,\sigma,\Lfonly)~\uplus
        \mathcal{L}(e_2,\sigma,\Lfonly)~\uplus\\
        &[\,\mathit{\epath} \mapsto  x.\clazy \sigma]
\end{align*}
The rules for  \CDR, \PRIM\ and \NULLQ\ are also  modified in a manner
similar to \CAR.  %% The definitions  of \bcar\, \bcdr\ and \clazy\ induce
%% the following relation $\hookrightarrow$ over sets of access paths:
%% \begin{align*}
%%   &\sigma_1\bcar\sigma_2  \hookrightarrow
%%   \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \acar\sigma_2',\\
%% &\sigma_1\bcdr\sigma_2  \hookrightarrow
%%   \sigma_1\sigma_2'\mbox{ where } \sigma_2 =
%%   \acdr\sigma_2',~\mbox{and}\\
%% &\sigma_1\clazy \sigma_2 \hookrightarrow \sigma_1, \mbox{ if }
%% \sigma_2 \ne
%%   \emptyset
%% \end{align*}
The reflexive  transitive closure of  $\hookrightarrow$ will
be   denoted    as   $\stackrel{*}{\hookrightarrow}$.    The
following proposition relates  the original and the modified
liveness rules.

>>>>

Consider the  $\mathcal{L}$-rules for {\LET}, {\SIF},  and {\SRETURN}.
The expression  $(\SRETURN~\psi:x)$ with  a demand $\sigma$  forces an
evaluation  of  $x$ and  hence  a  new map  $[\mathit{\epath}  \mapsto
  \{x.\sigma\}]  $ needs  to be  created,  where $\epath$  is a  fresh
evaluation path  terminating in  $\psi$.  The  map for  the expression
$(\SIF~\psi:x~e_1~e_2)$ is a pointwise-union of  the maps of $e_1$ and
$e_2$. In addition, since the condition also triggers an evaluation of
$x$, the map $[\mathit{\epath} \mapsto \{x.\epsilon\}]$ is created and
added  to the  union.  Also  notice a  consequences of  laziness---the
entire  expression including  the condition  is not  evaluated if  the
demand on it is $\emptyset$.  This  results in the empty map, which is
denoted by $\emptyset_{\cal{M}}$.

 The liveness  of {\LET} has  an elegant formulation.   Recollect that
 our language permits lazy {\LET}s whose definitions can be recursive,
 i.e.  in the expression $\LET~x \leftarrow~s~\IN~e$, $x$ can occur in
 $s$.   We  model  the  definition  $x  \leftarrow~s$  as  a  possibly
 recursive function (named $\mathit{f}$ in the rule) which captures an
 arbitrary number of ``unrollings'' of the definition\footnote{The $*$
   in  the definition  of ${\mathit  f}$ represents  non-deterministic
   choice.  During liveness analysis it  can be assumed to generate no
   liveness.}.  We  pass to  the demand  transformer $\Lfonly_{\mathit
   f}$ of this function the demand  arising out of the liveness of $x$
 at the beginning of $e$. This  gives the liveness of the variables of
 $s$.

@@

We make three observations: firstly the rule ({\sc live-define}) has a
least  solution  as  $\mathcal{L}(\cdot)$  is monotonic  in  $\sigma$;
secondly  that  ({\sc  live-define})   resembles  the  rule  for  type
inference of mutually recursive  function definitions, and thirdly the
asymmetry of  demand and liveness (compared to  post- and pre-liveness
classically) is due to the functional formulation here.
}

@@

The  solution  is  to  also   treat  the  operations  mentioned  above
symbolically.  We  introduce three new symbols:  \bcar, \bcdr, \clazy.
These symbols are defined as a relation $\hookrightarrow$ between sets
of access paths.  First, \bcar\ and \bcdr\ are defined as:

\begin{align*}
  &\bcar\sigma \hookrightarrow \sigma' \mbox{ where } \sigma' = \{\alpha \mid \acar\alpha \in \sigma\}\\
  &\bcdr\sigma \hookrightarrow \sigma' \mbox{ where } \sigma' = \{\alpha \mid \acdr\alpha \in \sigma\}
\end{align*}
Thus \bcar\ selects those entries in $\sigma$ that have leading \acar, and removes the leading \acar\ from the.
The symbol \clazy\ reduces the set of strings following it to a set containing only $\epsilon$. It filters out, however, the empty set of strings.
\begin{align*}
  \clazy\sigma \hookrightarrow & \left\{ 
  \begin{array}{ll}
    \emptyset&\mbox{if}~\sigma = \emptyset\\
    \{\epsilon\} & \mbox{otherwise}
  \end{array}\right.
\end{align*}
We can  now rewrite the \CONS\  and the \CAR\  rules of $\mathit{ref}$
as:
\begin{align*}
&\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
= x.\bcar\sigma \cup y.\bcdr\sigma  \label{eqn:mod-cons},~
\mbox{and} \\
&\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          =   x.\clazy\sigma \cup x.0\sigma
\end{align*}
and the \Lfunonly\ rule
for \SIF\ as:
\begin{align*}
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) =
                    &\mathcal{L}(e_1,\sigma,\Lfonly)~\uplus
        \mathcal{L}(e_2,\sigma,\Lfonly)~\uplus\\
        &[\,\mathit{\epath} \mapsto  x.\clazy \sigma]
\end{align*}
The rules for  \CDR, \PRIM\ and \NULLQ\ are also  modified in a manner
similar to \CAR.  %% The definitions  of \bcar\, \bcdr\ and \clazy\ induce
%% the following relation $\hookrightarrow$ over sets of access paths:
%% \begin{align*}
%%   &\sigma_1\bcar\sigma_2  \hookrightarrow
%%   \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \acar\sigma_2',\\
%% &\sigma_1\bcdr\sigma_2  \hookrightarrow
%%   \sigma_1\sigma_2'\mbox{ where } \sigma_2 =
%%   \acdr\sigma_2',~\mbox{and}\\
%% &\sigma_1\clazy \sigma_2 \hookrightarrow \sigma_1, \mbox{ if }
%% \sigma_2 \ne
%%   \emptyset
%% \end{align*}
When  there  are  multiple  occurences of  \bcar,  \bcdr  and  \clazy,
$\hookrightarrow$  is  applied  from  right to  left.   The  reflexive
transitive   closure  of   $\hookrightarrow$   will   be  denoted   as
$\stackrel{*}{\hookrightarrow}$.   The  following proposition  relates
the original and the modified liveness rules.
